---
title: "Control Structures"
author: "Varun Boodram"
date: "September 24, 2014"
output: html_document
---
### If-else construction

The format for an if else loop is

if(condition){  
        do something  
}else if {  
        do another thing  
}else {  
        do another thing  
}
```{r}
sety<-function(x){
        if(x>3){
                10
        }else {
                0
        }
}
y<-sety(4); y
```
Rather than write a function, y can be set to the entire if-else construct
```{r}
x<-4
y<-if(x>3){
                10
        }else {
                0
        }
y
```
Multiple condtions use & and |
```{r}
sety2<-function(x){
        if(x>3 & x<=6){
                y<-2
        }else{
                y<-1
        }
}
y<-sety2(6); y
y<-sety2(10); y
```
```{r}
setanothery<-function(x){
        if (x==2 | x==4 | x ==6 | x==8 |x==10 ){
                print("x is even")
        }else if (x==1 | x==3 | x ==5 | x==7 |x==9 ){
                     print("x is odd")
             }else{
                     print("enter x between 1 and 10")
                     }

}
setanothery(3)
setanothery(8)
setanothery(100)
```
Conditional element selection can also be acheived with ifelse, but readability diminishes with complexity; probably only suitable for only a couple of conditions
```{r}
# ifelse(test, yes, no)
x<-1:10
ifelse(test = x<5|x>8, yes =  0, no =  x)
ifelse(test = x>= 4 & x<=8, yes = ifelse(test = x!=7, yes = 0, no = 7 ), no = x)
# original sety function
x<-c(3, 8)
ifelse(test = x > 3, yes = 10, no = 0)

# setanothery
x<-c(3, 8, 100)
ifelse(test = x==2 | x==4 | x ==6 | x==8 |x==10, yes = print("x is even"), no = ifelse(test = x==1 | x==3 | x ==5 | x==7 |x==9 , yes = print("x is odd"), no = print("enter x between 1 and 10")))
```
### While loop

The format of a while loop is 

while(condition is true){  
        do something  
}
```{r}
count<-0
while(count<10){
        print(count)
        count<-count+1
}
```
### Repeat loop

Repeat loops initiate an infinite loop; this is pointless, unless break is called at some point, which exits the loop. For loops are prefereable
```{r}
c<-5
repeat{
        if(c<0 | c>8){break}
        set.seed(5)
        coin<-rbinom(n = 1, size = 1, prob = 0.5)
        if(coin<0.5){
                c<-c+1
                print(c)
        }else{
                c<-c-1
                print(c)
                }
}
```  
### for loop and loop functions (the apply family)
The for loop is the most common type of operator that will be used. The format is 

for (a number of times){  
        do something  
}
```{r}
x<-sample(1:5); x
for (i in 1:5){
        print(x[i])
}
for (i in seq_along(x)){
        print(x[i])
}
```
for loops can be nested
```{r}
x<-1:10
dim(x)<-c(5,2)
for (i in seq_len(nrow(x))){ #sequence along the length of nrow
        for (j in seq_len(ncol(x))){
                print(x[i,j])
        }
}
```
Deep nesting can result in code that is difficult to understand

Loop functions execute a loop using just one line of code. 
 
*lapply* loops over a **list** and evaluates a function on each element
```{r}
x<-sample(1:5); x
lapply(X = x, function(x) x)
```
The first argument of lapply is a **list**; the second is the function that you want to apply to each element of that list. If the first argument is not a list, it is coerced to be, which is why the above output was given as a list. 
```{r}
class(lapply(X = x, function(x) x)) #this is an anonymous function, cf, eg: mean, sum, etc
```
lapply returns something for every value in the input list.
```{r}
lapply(X = 1:4, FUN = rnorm) # the rnorm function will work with defaults
lapply(X = 1:4, FUN = rnorm, mean = 20 ,sd = 5) # defaults are changed outside of the brackets (just delete brackets after specifying vals)
```
sapply is a variant of lapply that simplifies the result, if it is possible
```{r}
sapply(X = x, function(x) x) #possible to simplify
sapply(X = 1:4, FUN = rnorm, mean = 20 ,sd = 5) #not possible to simplify
```
*apply* evaulates a function over the margins of an array. A margin is a fancy word meaning "by row" or "by column". Recall that an array is a list of matrices
```{r}
x<-1:20; dim(x)<-c(2,2,5); x
```
apply operates on this list by applying a function to margin = 1 (by rows), or margin =2 (by columns)
```{r}
# a matrix is a two dimensional array
y<-matrix(data = 1:10, nrow = 5, ncol = 2); y
apply(X=y, MARGIN = 2, FUN = sum)
#check
c(sum(1:5), sum(6:10))
# the previously constructed x is a 3 dimensional array
apply(X = x, MARGIN = 1, FUN = sum)
#check
c(sum(1,3,5,7,9,11,13,15,17, 19), sum(2,4,6,8,10,12,14,16,18,20))
# mean of all matrices in the array (collapsing the third dimension)
apply(X = x, MARGIN = c(1,2), FUN = mean)
# mean of the second and third matrices in the array
apply(X = x[,,2:3], MARGIN = c(1,2), FUN = mean)
# sum of the elements of each matrix in the array
apply(X = x, MARGIN = 3, FUN = sum)
```
**tapply** applies a function to a subset of a vector. The first argument is a vector (or vector-like object), and the second argument is a factor vector of the same length as the first object, that identifies to which group each element of the original object belongs. 
In the "data types notes", there is a bit on creating a factor variable using factor(). Factor levels can also be generated with gl() (**g**enerate factor **l**evels)
```{r}
gl(n = 5, k = 2, length = 10) # a 10- element (length =) factor vector with 5 (n =) levels, each repeated twice
gl(n = 4, k = 5, length = 20)
x<-1:20; x
f<-gl(n = 5, k = 4, length = 20); f
tapply(X = x, INDEX = f, FUN = mean, simplify = T)
tapply(X = x, INDEX = f, FUN = summary, simplify = T)
```
**mapply** is a multivariate version of lapply and sapply. It is useful for vectorization. Think about it this way: the functions used so far (mean, sum, etc), take a single argument. Suppose we want to use a function that takes two arguments. mapply can be used here
```{r}
list(a=rep(1,4), b=rep(2,3), c=rep(3,2), d=rep(4, 1))
```
It's easier to use mapply with two inputs: one for the first argument of each rep, and one for the second
```{r}
list<-mapply(FUN = rep, 1:4, 4:1, SIMPLIFY = T)
names(list)<-c("a","b","c","d"); list
```
**split**  
a common idiom is 

lapply(X = split(somthing), FUN = some function)  
split does what it sounds like: it splits a list or data frame by factor
```{r}
split(x = x, f = f)
split(x = mtcars$mpg, f = mtcars$gear)
```
Note that tapply does the same thing as the idiom. 
```{r}
sapply(X = split(x,f), FUN = sum)
tapply(X = x, INDEX = f, FUN = sum)
```
The advantage of split is that it can be used to split much more complex objects. 
```{r}
head(airquality)
s<-split(x = airquality, f = airquality$Month)
sapply(X = s, function(x) colMeans(x[,c("Ozone", "Solar.R", "Wind")], na.rm = T))
```
split can even split a data frame on multiple levels
```{r}
head(mtcars)
```
Suppose that we are interested in how miles per gallon varies with the numbers of gears, cylinders, and carburetors. Convert each of these to a factor:
```{r}
f1<-as.factor(mtcars$gear)
f2<-as.factor(mtcars$cyl)
f3<-as.factor(mtcars$carb)
#view interactions
interaction(f1,f2,f3)
# obtain a ridiculously fine grain of detail
split(x = mtcars, f = list(f1,f2,f3))
```