---
title: "Statistical Inference Notes"
author: "Varun Boodram"
date: "October 1, 2014"
output: html_document
---

The goal of statistical inference is to infer facts about a population from a sample of that population

## Basic Probability
Given a random experiment, a probability measure is a **population quantity** that summarizes the randomness

* in a die roll experiment, a probability measure is an intrinsic property of the die
* probability does not exist in the data that we have, but as a conceptual thing that exists in the population. 

### Axioms
If $S$ is the sample space, 

* $P(S)=1$; $P(S^c)=P(\phi))=0$ i.e.: something must occur; the probability of nothing occurring is 0
* If $A\in S, 0\leq P(A)\leq1$
* If $A \cap B \neq 0, P(A \cup B) =P(A)+P(B)-P(A\cap B)$


### Examples
A **probability distribution** lists all possible outcomes in the sample space and the probabilities with which they occur. 

eg: Let $X$ be the result of a coin flip, where $X=0$ represents tails. Let $P(\theta)$ be the probability of a tail. 
$$P(x)=\theta^{x}(1-\theta)^{(1-x)}, x= 0, 1$$
This is useful for modeling the prevalence of something (say, high blood pressure). 

eg: Suppose the proportion of help calls that get addressed in a random day by a help line is given by 
$$
f(x) = \left\{
        \begin{array}{ll}
            2x & \quad 0 < x < 1 \\
            0 & \quad else
        \end{array}
    \right.
$$

```{r}
x<-c(-.5, 0, 0.5, 1, 1.000001, 1.5)
y<-sapply(X = x, function(x) ifelse(test = 0<x & x<=1,yes = 2*x, no = 0))
plot(x, y, lwd=3, frame=F, type="l")
```

What is the probability that 75% or fewer calls get addressed in a randomly sampled day from this population?

```{r}
height<-2*0.75
prob<-height*0.75/2; prob
```

The above density is a special case of what is called a *beta distribution*. The letter *p* in front of a density gives the probability
```{r}
pbeta(q = 0.75, shape1 = 2, shape2 = 1)
```
p+density name returns the CDF of a random variable $X$, $$F(x)=P(X\leq x)$$
The **survival function** $S(x)=1-F(x)=P(X>x)$

### Quantiles
The $\alpha$th **quantile** of a distribution with the function $F$ is the point $x_{\alpha}$ such that $$F(x_{\alpha})=\alpha$$
q+density name gives the qth quantile
```{r}
median<-qbeta(p = 0.5, shape1 = 2, shape2 = 1); median
```

### Conditional Probablilty
If $P(B)> 0$, $P(A\vert B)\overset{\Delta}{=}\frac{P(A \cap B)}{P(B)}$  
If $P(A \vert B)=P(A)$, $A$ and $B$ are independent.   
Bayes' Rule
$$P(B\vert A)=\frac{P(A\vert B)P(B)}{P(A\vert B)P(B)+P(A\vert B^c)P(B^c)}$$

eg: 29.5 % of a population is infected with HIV.  An HIV test is 99.7% accurate for people carrying HIV, and 92.6 % accurate for those who don't carry HIV. If an individual tests positive, what is the probability that he carries HIV? 

solution steps

1. Identify what you are trying to figure out
$$P(\text{carries HIV}\vert +)$$
2. Draw tree diagram
```{r}
require(openintro)
treeDiag(c('Carries HIV?','Tests Positive?'),
        c(.295,(1-.295)), list(c(.997,(1-.997)), c((1-.926),.926)))
```
3. $P(\text{carries HIV}\vert +)= \frac{P(\text{carries HIV} \cap +)}{P(+)}=\frac{0.2941}{0.2941+0.0522}$