---
title: "Subsetting"
author: "Varun Boodram"
date: "September 23, 2014"
output: html_document
---

Depending on the class of the R object, different operators will have to be used to extract subsets of the object  

### Numerical subsetting

* [ ] always returns an object of the **same** class as the original. 

eg: vectors
```{r}
x<--(1:10); x
x[3] # third element of the vector
c(class(x), class(x[3]))
```
eg: lists
```{r}
l<-list(foo=1:4, bar=matrix(data = 1:4,nrow = 2, ncol = 2, byrow = T))
l
l[2] # second element of the list
c(class(l), class(l[2]))
```
Note that l[2] is a list of one element, called "bar", which is a matrix. If the elements of a list are named (as above), the name of an element index can be used to return a list. This is preferrable to use, as you don't have to remember where the element is in a list
```{r}
l["bar"]
#for fun, change the names of the list
names(l)<-c("hello", "goodbye")
l["goodbye"]
```
Extracting multiple elements from a list *requires* the use of [ ]. 
```{r}
l2<-list(foo=1:4, bar=matrix(data = 1:4,nrow = 2, ncol = 2, byrow = T), norf = "hello")
l2[c(1,3)]
l2[c("bar","norf")]
```
eg: data frame
```{r}
head(iris[3])
c(class(iris), class(iris[3]))
```
Matrices are an exception to the rule, "[ ] always returns an object of the same class as the orginal object." Instead of a *matrix*, by default, subsetting results in a *vector*  
eg: matrix
```{r}
m<-1:10
dim(m)<-c(2, 5)
m[1,4] # row 1, column 4
c(class(m), class(m[1,4]))
```
This behaviour can be turned off by setting the drop argument to F
```{r}
m[1,4, drop =F]
c(class(m), class(m[1,4, drop =F]))
```
* [[ ]] is used to extract *a single element* of a **list** or a **data frame**. This element may or may not be of the same class as the subsetted object

eg: lists
```{r}
names(l)<-c("foo", "bar")
l[[2]]
c(class(l), class(l[[2]]))
```
The difference between l[2] and l[[2]] is that l[2] returns the one-element *list*, l[[2]] returns the object(s) in that list.   
If the numerical index has a name, then the name can be used instead of the numerical index
```{r}
l[["foo"]]
l[["bar"]]
```
Suppose you want to access the third element of the vector named foo. This is accomplished with *pairs of [[ ]]*
```{r}
l[[1]][[3]]
```
An alternative syntax uses a vector to access the element
```{r}
c(l[[c(1,3)]], l[[c(2, 3)]])
```
eg: data frame 
```{r}
head(iris[[1]])
c(class(iris), class(iris[[1]]))
```
* $ is used to extract elements of a **list** or a **data frame** by name. The extracted element may or may not be of the same class as the original object. 
eg: list
```{r}
l$foo
l$bar
c(class(l), class(l$foo), class(l$bar))
```
$ can *only* be used with literal names. 
```{r}
myname<-"foo"
l$myname
```
Because myname is *not* a literal name, l$myname is meaningless. To obtain the object(s) stored under the name myname, [[ ]] are required
```{r}
l[[myname]]
```
Thus,  *computed* indices are accessed with [[ ]]   

### Logical subsetting
This is used when you want to subset by a *condition*. 
eg: vector. The condition goes inside the brackets
```{r}
x<-c("a","b", "c", "c", "d", "a")
x[x>"a"]
```
Logical indexing can be used to create a logical vector
```{r}
u<-x>"a"; u
```
This logical vector can now be used as a condition
```{r}
x[u]
```
This princliple can be used to remove missing values. 
```{r}
x<-c(1:8, rep(NA, 8))
x<-sample(x); x
keep<-!is.na(x); keep
x[keep]
```
The complete.cases() function can be invoked when you want to perform a similar operation on more than one vector, *each of which have the same length*
```{r}
x<-c(1:8, rep(NA, 8)); x<-sample(x); x
y<-c(letters[1:8], rep(NA, 8)); y<-sample(y); y
keep<-complete.cases(x,y)
rbind(x[keep], y[keep])
```

Logical subsetting works on the same principles as numerical subsetting
```{r}
l
l[names(l)=="bar"]
l[names(l)>"bar"]
l2
l2[names(l2)>"bar" | names(l2)=="norf"]
```